Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-07-19 review_article.Rmd
Included table of contents

---
title: "Cartogram Mapping and its Application to Cancer Data Visualisation"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  html_document: default
fontsize: 11pt
link-citations: true
fig_caption: yes
biblio-style: vancouver
---

@@

Utilising maps to present statistics has been widely used for centuries.
These maps connec data and/or statistics to the geographical representation of areas that are generally familiar to the audience. However, it is not enough to be recognisable, if the meaning of the statistic cannot be seen. This is particularly relevant to cancer statistics, as cancer outcomes directly relate to the people living within a geographical area. In situations where the people are of interest, not the land they live on, it is reasonable to explore views that enhance the communication of the cancer statistics. This has spurred innovations over the previous centuries to enhance the shapes and maps presented to effectively communicate cancer outcomes, and health outcomes more braodly.

@@

@@

<!-- what is a choropleth map -->
A choropleth map is used to show the spatial variability of measurements, with usage dating back to the 1800s. They are constructed by drawing the given geographic or political boundaries, and filling the polygon with colour to represent values of a measured variable [@EI]. Early versions used symbols or patterns instead of colour. @CPISACA discusses the use of choropleth maps for visualising cancer data, and @DMAHP gives an overview of the development of the use of choropleth maps for displaying disease data. A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns across a familiar geographic form, that might show trends in disease occurrence, or even localised outbreaks. Figure 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.

<!--
@GOINO major geographic storage device; map reading and map interpretation have been the basic methods of retrieval and analysis.
Geography’s interests are in spatial distributions and associations, in areal differentiation
and in spatial processes. All of these interests demand information that is geographically
ordered. Geographically ordered data are those data which have their locational
coordinates preserved.
-->

@@

<!-- psychology -->
Utilising the familiar state shapes can make a map intuitive to read [@CIBMUK], and allow viewers to visually infer the spatial relationships in the data.
The familiarity of the geography is a worthy consideration when presenting results of spatial analysis. Just as geographers are no longer the only creators of maps, @CPISACA suggests the audience of spatial health data analysis has extended beyond researchers to the public, policymakers and the media.
However, while the areas are recognisable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, different population densities and subject to changes over time. 
The different sizes of administrative areas can attract attention to the shades of the unpopulated, large physical areas [@EI]. Choropleths can inhibit visual inference when presenting human related statistics as the display may draw attention from the 'potentially more important results in the more populous communities' [@SE].

<!--
Certain limitations to the standard graphic techniques have also become apparent, however.
They are slow, and do not permit instant retrieval of information. They are untrustworthy when interpretations of spatial association are needed; there are both severe physiological and psychological constraints on our abilities to assess correlations of spatial distributions visually.

These opportunities will be lost if locational coding is not inserted into the data referencing system as such banks are built, however.

of some element, for example population density, or purchasing power, or agricultural fertility; reality, of course, is not like this. the background variable assumed to be uniform (such as population density) can be adjusted so that it behaves as if uniformity were the case
-->

<!--
Urbanisation has pushed people closer together,
-in workplaces and in homes. 
-this issue is not with the size of the administrative areas,
- the increasing urbanisation is contributing to the changing density of population
-->

<!-- Displaying cancer data on choropleths -->
In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality. 
Almost 100 years of cancer maps have explored the United States and the United Kingdom with increased effectiveness in presentation of unbiased rates. The increasing development and use of disease maps can be attributed to the availability of geographic information system software [@SE]. The data collection methods of cancer mortality rates across regions, and administrative control within regions lends itself choropleth visualisation. Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the proportion of old people. Howe [-@HEDP] describes Stocks development of the standardised mortality ratios through the 1930s. The choropleth maps used presented levels of cancer via hatchings on a black and white scale.

<!--
Computing mapping methods improved with access to necessary information: locations and associated data points. @GOINO discusses progression in computer mapping methods with the availabilty of census data with associated locations.
-->

<!-- Usage and availability -->
@DMAHP cites Cruickshank's (1947) discussion of using visuals as 'formal statistical assessment of the spatial pattern' as a major advancement.
@CD considers cartographic data analysis in practice. Presenting a clear guide for preparing cartograms for use with both qualitative and quantitative information.
The authors recognise that the creation of cartograms was largely the work of professional cartographers until the innovation of geographical information systems that welcomed map users as map creators. These systems are utilised depending on ‘the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994)’. @MACM also recognises the intuition that can be derived from choropleths must be 'validated by rigorous statistical analyses.'

@@

A grey literature review conducted by (ref) identified 33 publically available cancer atlases published on the internet between 01/01/2010 to 11/11/2015.

@@

<!-- Going from choropleth to cartogram -->

@@

 @CTTMB argues that using choropleth maps for population characteristics requires graphic distortions when the population concerned varies greatly in density. 
When implementing a distortion of the geographical shape according to population, an area cartogram [@NAC], or iso-demographic map is the result.
When visualising population statistics @ACTUC considers this equitable representation design 'more socially just, giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. @CTTMB suggest that spatial socio-economic data is best presented on a cartogram for urban areas.
@HEDP agrees that 'cancer occurs in people, not in geographical areas' and the map bases of population reflect this and avoid allocating 'undue prominence' to rural areas. @MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities. Using health estimates from large areas can prevent drawing attention to the inequalities.

@@

A contiguous cartogram allows the map space to highlight the distribution of the variable. These cartograms maintain connectivity of the map regions while areas are resized according to a statistic, this often must occur at the expense of the shape [@CBATCC,@NAC].
From a computer graphics perspective, @ACA believe it is a problem of 'map deformation' to account for the value assigned to each area, they provide three methods for creating value-by-area cartograms.
Examples include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [-@ACTUC], Radial Expansion Method of Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and Tikunov’s Line Integral Method, Constraint-Based Method (Kocmoud and House) [-@CBATCC].
- compare to their

@@

@@

>>>>

---
title: "Cartogram Mapping and its Application to Cancer Data Visualisation"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  html_document: default
fontsize: 11pt
link-citations: true
fig_caption: yes
yolo: true
biblio-style: vancouver
---

@@

Utilising maps to present statistics has been widely used for centuries.
Connecting the data to the geographical representation of areas that are already familiar.
It is not enough to be recognisable, if the value of the statistic cannot be seen. Cancer outcomes directly relate to the people living within a geographical area. In situations where the people are of interest, not the land they live on, it is reasonable to explore views that enhance the communication of the cancer outcomes. This has spurred innovations over the previous centuries to enhance the shapes and maps presented to effectively communicate cancer outcomes.

@@

<!--
Disease maps can rapidly highlight geographical
patterns that would otherwise be hidden in a
table, and are commonly used for hypothesis
generating and as the first step in exploratory
spatial data analysis. @SE
Geography is concerned with the identification and
explanation of spatial structure, pattern, and process,
and with the analysis and explanation of the links
between humans and the environment (2). @SAMGIS

By providing a visual representation of
cancer outcomes, geographic patterns of disease are able
to be identified and effectively addressed
-->

@@

<!-- what is a choropleth map -->
A choropleth map is used to show the spatial relationship, or variability, of measurements. Created as a storage device which preserves locations for geographically ordered data, with usage dating back to the 1800s [@GOINO]. They are constructed by drawing the given geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI]. Early versions used symbols or patterns instead of colour. @CPISACA discusses the use of choropleth maps for visualising cancer data, and @DMAHP gives an overview of the development of the use of choropleth maps for displaying disease data. A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns across a familiar geographic form. Figure 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.

<!--
Thematic maps, on the other hand, do select some geographical concept or arrangement and attempt to focus attention entirely on the spatial characteristics of the entire distribution and upon its relation to other geographical factors.
-->

@@

<!-- psychology -->
Utilising the familiar state shapes can make maps intuitive to read [@CIBMUK], and allow viewers to visually infer the spatial relationships in the data.
The value of the familiarity of the geography is worth consideration when presenting results of spatial analysis. 
Just as geographers are no longer the only creators of maps, @CPISACA suggests the audience of spatial health data analysis has extended beyond researchers to the public, policymakers and the media.
However, while the areas are recognisable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, different population densities and subject to changes over time. 
The different sizes of administrative areas can attract undue attention to the shades of the unpopulated, large physical areas [@EI]. @GOINO believes there are "severe physiological and psychological constraints on our abilities to assess correlations of spatial distributions visually".
As the display may draw attention from the 'potentially more important results in the more populous communities' [@SE], choropleths can inhibit visual inference when presenting human related statistics.

<!--
Urbanisation has pushed people closer together,
-in workplaces and in homes. 
-this issue is not with the size of the administrative areas,
- the increasing urbanisation is contributing to the changing density of population

A particular spatial structure includes
the individuals affected and how they are connected in
communities, as well as the dynamics of these communities
and their organization into larger units. @SAMGIS
-->

<!-- Displaying cancer data on choropleths -->

In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality. 
Almost 100 years of cancer maps have explored the United States and the United Kingdom with increased effectiveness in the presentation of unbiased rates. The increasing development and use of disease maps can be attributed to the availability of location data and associated data points [@GOINO], and geographic information system software [@SE]. Choropleth visualisation suits the data collection methods of census data and cancer mortality rates across regions, and administrative control within regions.

Mortality rates are now often presented as relative rates of risk across the population, and age adjusted. @HEDP describes Stocks' development of the standardised mortality ratios through the 1930s. The choropleth maps used presented levels of cancer via hatchings on a black and white scale.

<!--
@DMAHP cites Cruickshank's (1947) discussion of using visuals as 'formal statistical assessment of the spatial pattern' as a major advancement.
@MACM also recognises the intuition that can be derived from choropleths must be 'validated by rigorous statistical analyses.'

@CD considers cartographic data analysis in practice. Presenting a clear guide for preparing cartograms for use with both qualitative and quantitative information.
The authors recognise that the creation of cartograms was largely the work of professional cartographers until the innovation of geographical information systems that welcomed map users as map creators. These systems are utilised depending on ‘the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994)’. 
-->

<!--
of some element, for example population density, or purchasing power, or agricultural fertility; reality, of course, is not like this. the background variable assumed to be uniform (such as population density) can be adjusted so that it behaves as if uniformity were the case
@GOINO
-->

@@

A grey literature review conducted by (ref) identified 33 Cancer Atlases publically available cancer atlases available on the internet and published between 01/01/2010 to 11/11/2015.

@@

<!-- Going from choropleth to cartogram 
A cartogram is a map purposely distorted so that its spatial properties represent quantities
not directly associated with position on the globe. As thematic maps, cartograms emphasize
the distribution of a variable by changing the area (or lengths) of objects on the map.
Our approach maintains continuity, and preserves
many local features of cartographic lines that provide visual clues to the identity of the distorted
objects.
@ACCAC
-->

@@

<!--
@SE Researchers mapping spatial patterns of disease
for small areas are confronted with a number of
challenges, including cartographic design, the
modifiable area unit problem, and also the “small
number” problem, where some places have no,
or few, disease events and/or a low population at
risk of contracting the disease.

@TAAM variations in population density may obscure geographic patterns
This factor may be eliminated by making the map area proportional to population rather than to actual land area.
Such maps, or what have been termed "population-by-area cartograms," are well known in the
cartographic literature
-->
Using choropleth maps for population characteristics requires graphic distortions when the population concerned varies greatly in density [@CTTMB].
When implementing a distortion of the geographical shape according to population, an area cartogram [@NAC], or iso-demographic map is the result.
When visualising population statistics @ACTUC considers this equitable representation design 'more socially just, giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. @CTTMB suggest that spatial socio-economic data is best presented on a cartogram for urban areas.
@HEDP agrees that 'cancer occurs in people, not in geographical areas' and the map bases of population reflect this and avoid allocating 'undue prominence' to rural areas. @MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities. Using health estimates from large areas can prevent drawing attention to the inequalities.

@@

A contiguous cartogram allows the map space to highlight the distribution of the variable. These cartograms maintain connectivity of the map regions while areas are resized according to a statistic, this often must occur at the expense of the shape [@CBATCC,@NAC, @TAAM].
From a computer graphics perspective, @ACA believe it is a problem of 'map deformation' to account for the value assigned to each area, they provide three methods for creating value-by-area cartograms.
Examples include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [-@ACTUC], Radial Expansion Method of Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and Tikunov’s Line Integral Method, Constraint-Based Method (Kocmoud and House) [-@CBATCC].

<!--
Medical application: TAAM 1965
Actual shapes of counties were sacrificed to the extent necessary to retain as far as possible, common boundaries, relative locations, and the approximate contour of the State.

There is no unique solution to the topologic
problem of preparing a population-by-area
cartogram for a given population and its jurisdictional
boundaries. Although the map areas
of the various regions are fixed by their popu¬
lations, their shapes and relative locations are
not completely determined.

Other characteristics of the population
may also serve as the base for adjustment; for
example, sex, race, parity, socioeconomic level,
or iminune status.

@RSCW
helps to correct misconceptions. Pupils are likely to over-estimate regions as Canada, Siberia, Africa, and our own West just because they loom up large on maps. The simple picture which the cartogram gives is easily comprehended and remembered and
will be stimulating in making comparisons.

@NISCC
two advantages of the cartogram: (1) new relationships are often revealed visually, and ( 2 ) it is often more honest to use “other” space on the base map than land area. Finally, and most important for the present discussion, they suggest that the ability to preserve shape of the statistical units drops with an increase in the number of these units, and in situations where the difference is great between the smallest and largest values mapped.
Until we know how the map reader compares it to an ordinary thematic map, we are experimenting with a graphic device that may be reducing (or at least hindering) the flow of information from the map to the map reader.
-->

@@

<!--
-simple shapes of the circle, square, and triangle are relatively easily recognized.
-sacrifice communication efficiency;
-There is evidence, too, that angular shapes are more easily seeu than ones with smooth outlines. Shapes that contain the qualities of simplicity and reg,ularity are more readily perceived than others that do not possess these qualities.
-Finally, psychologists tell us that shapes having familiarity are more quickly
recognized than nonsense shapes or those that have no meaning.
-The cartographer can oiily efficiently communicate with this kind of map if he has preserved particular elements of the original outline and incorporated them in the new shape.
-
@NISCC 
-->

@@

<!-- 
TVSSS parliamentary cartogram where each constituency is represented by an oval of equal size. Many ovals overlap because the cartogram is still in the process of being made. Each
-->
